Periodicity of Geodesics on the Necker Cube Surface

### 2018-01-04 backup.tex
010318

@@

\subsection{Discussion of Results}
Our initial experiments supported the theory that there would be a strong correlation between a choice of initial trajectory angle and the dynamical properties of a geodesic traveling on the surface. [throw in some trajectories in the appendix] A surface composed of infinitely many cubes is endowed with a flat metric, i.e. every neighborhood is locally isometric to $\RR^2$. Via the parallel transport of a unit vector tangent to the surface in the direction of the geodesic, we view its image as a sequence of straight lines on the (smooth) faces of the surface that tightly wrap around its sharp edges (Fig 1).
\\
\begin{wrapfigure}{l}{2in.}
\includesvg{vectorplane}
\end{wrapfigure}
\noindent Let $u_0$ be a point on the surface that is contained in the interior of a face and consider a tangent unit vector $v\in\RR^3$ based at point $u_0$. Each face is parallel to a $2$-dimensional subspace of $\RR^3$ spanned by exactly two basis vectors, so one component of $v$ is always 0. Thus $v$ is is isometric to $v_0\in\RR^2$.
\\\\\\
Denote the Necker cube surface $\bUs$. The angle $v_0$ makes relative to a choice of basis is the \emph{initial trajectory angle} of a \emph{unit-speed geodesic} based at $u_0$ on the surface, $\Phi_t:\RR\rightarrow\bUs$. The initial trajectory angle is said to be $\emph{rational}$ if there is some $k\in\RR$ such that $kv_0$'s components are relatively prime integers. A rational initial trajectory angle in $\RR^2$ falls into one of two categories:
\begin{Def}
Let $v_0$ be a rational unit vector of the form $\frac{1}{k}(x,y)\in\RR^2$ with $x,y\in\mathbb{Z}$ and $k=\sqrt{x^2+y^2}\in\mathbb{R}$. We say $kv_0$ is an \textbf{odd-odd} vector if $x,y$ are relatively prime and both odd. We denote the \textbf{set of all odd-odd directions} by $\mathcal{O}$. We say that $kv_0$ is an \textbf{even-odd} vector if $x,y$ are relatively prime and of opposite parity. We denote the \textbf{set of all even-odd directions} by $\mathcal{E}$.
\end{Def}
\noindent Let $\cS=\cO\sqcup\cE$ be the disjoint union of the two. This paper aims to depict the relationship between initial trajectory angles and dynamical properties of a geodesic on the surface. Denote $\bUs$'s translational isometries by $\text{Trans}(\bUs)$. Recall that a geodesic on the surface is \emph{periodic} if there exists some $T>0$ such that $\Phi(t+T)=\Phi(t)$ for all $t\in\RR$ and a geodesic is \emph{drift-periodic} if there exists some $T'>0$ and non-trivial $p\in \text{Trans}(\bUs)$ such that $\Phi(t+T')=p(\Phi(t))$ for all $t\in\RR$. In either case we call this value the \emph{period} of $\Phi$ if it is the \emph{smallest} possible value that satisfies this definition. We additionally require that at no point does $\Phi$ encounter a cone singularity of the surface. Otherwise the geodesic behavior is indeterminable. We call $\Phi$ \emph{non-singular} if this holds.

@@

\begin{thm*}
(Dynamical Classification of Geodesics on the Necker Cube Surface) \\Let $\Phi_t:\RR\rightarrow\bUs$ be a non-singular unit-speed geodesic on the Necker cube surface such that $\Phi(0)=u_0$ and $\Phi'(0)=v\in\RR^3$, isometric to $v_0\in\RR^2$. Let $\overline{u}_t\in\RR^3$ be $\Phi_t$'s coordinate in $\bUs$'s ambient space at time $t$. Let $k\in\RR$ such that $kv_0\in\cS$. Then the following is true:
\begin{enumerate}[label=(\roman*)]
\item $\Phi$ is periodic if $kv_0\in\cO$.
\item Suppose $\Phi$ is periodic. If $M (kv_0)=(1,1)$ then $\Phi$ has period $T=6\sqrt{2}||kv_0||$.
\item $\Phi$ is drift-periodic if $v_0\in\cE$. 
\item Suppose $\Phi$ is drift-periodic. If $M (kv_0)=(1,0)$ or $M (kv_0)=(0,1)$ then $\Phi$ has period $T=2||kv_0||$, and $d_E(\overline{u}_t,\overline{u}_{t+T})=||kv_0||\sqrt{2}$.
\end{enumerate}
\end{thm*}
\compav{Remember to prove that such an M exists.}

@@

We will call this ``half-cube" $\bG$. It is a genus one piecewise smooth surface homeomorphic to $\TT^2$. Every vertex has a cone angle of either $3\pi$ or $\frac{3\pi}{2}$. Let $\Sigma_{\bG}\subset\bG$ be the set of these three vertices. These are conical singularities of the surface. We denote the surface punctured at these points by $\bG\backslash\Sigma_{\bG} = \bGs$. The paths labeled $a,b,c,d$ are independent of one another and span $\bGs$'s first fundamental group, $\pi_1(\bGs)\cong \<a,b,c,d \>$, the free group of four generators. The unpunctured Necker cube surface $\bU$ is embedded in $\RR^3$ by tiling along $\bG$'s outer edges:

\begin{figure}[H]
\centering
\includesvg[width=4in]{neckercubepathsurface}
\end{figure}
From this diagram we see that $\bU$ is a universal cover of $\TT^2$. In this paper we are studying geodesic behavior on $\bU$ and in order to do so, it is necessary to determine how the parallel transport of a unit vector around a cone singularity acts on the unit tangent bundle of the surface. But to study the path monodromy we require that these loops (see $c,d$ in Fig 3) aren't contractible on the cube's sharp corners. Denote the countably infinite set of singularities of $\bU$ by $\Sigma_{\bU}$. The \emph{Necker cube surface} is the punctured surface $\bUs=\bU\backslash\Sigma_{\bU}$. Observe from the previous diagram that $\pi_1(\bUs)$ is the kernel of the following group homomorphism:

\begin{Def}
Denote $\varphi_1:\pi_1(\bGs)\rightarrow \ZZ^2$ as the group homomorphism such that $c,d\mapsto(0,0)$, $a\mapsto(1,0)$, and $b\mapsto(0,1)$.
\end{Def}

Take $\pi_1(\bUs)$ to be $\ker\varphi_1$. Then $\Delta_{\varphi_1}=\pi_1(\bGs)/\pi_1(\bUs)\cong\ZZ^2$ is the \emph{deck group} of the covering map $\varphi_1^*:\bUs\rightarrow\bGs$. Now $\bGs$ is unfolded and flattened in order to study monodromy as it transports a vector in $\RR^2$ along a path. Apply the following cutting and gluing operations to the surface: \\
\begin{figure}[H]
\centering
\includesvg[width=3in]{LTorus}
\end{figure}
\noindent This L-shaped torus is then taken apart and recovered as $2\times 2$ torus missing a unit square with the following identifications:\\
\begin{figure}[H]
\centering
\includesvg[width=2in]{BranchTorus}
\caption{$\bGs$ and images of paths after reconstruction. The outermost vertex is \emph{not} a puncture.}
\end{figure}

@@

\noindent Let $\aleph_{m,n}=2m+i2n\in\CC$ be the center of a removed square. Now let $\zeta_{m,n}(t)=(2m+t)+i(2n-\frac{1}{2})$ and 
$\xi_{m,n}(t)=(2m+t)+i(2n+\frac{1}{2})$ be paths on $\mathbf{P}$ that run along the edges of the squares, parameterized by $t\in (-\frac{1}{2},\frac{1}{2})$ and $(m,n)\in\ZZ^2$. $\bUs$ is recovered as a topological quotient $\mathbf{P}/\sim$, where $\sim$ is the minimal relation on $\mathbf{P}$:

\begin{equation}
\begin{split}
\zeta(t)\sim e^{i\frac{\pi}{2}}~\overline{\zeta(t)-\aleph}\\
\eta(t)\sim e^{i\frac{\pi}{2}}~\overline{\eta(t)-\aleph}
\label{eq:rel2}
\end{split}
\end{equation}
The translations, $\pm 2,\pm 2i$, on $\mathbf{P}$ are isometries that preserve these identifications and induce isometries of $\bUs$. The group generated by these horizontal and vertical translations form the \emph{group of Deck transformations}, $\text{Deck}(\varphi_1^*)$, that  is isomorphic to $\ZZ^2$:

@@

Identify 2 with $f_h$, and $2i$ with $f_v$. Let $x_0\in\bGs$ and fix a point in the fiber,  $u_0\in\varphi_1^*{-1}(x_0)\subset\bUs$. The action of $\ZZ^2$ on the fiber over $x_0$ is given by $(m,n)\cdot u_0=(f_h^m\circ f_v^n)(u_0)$. The distance between $u_0$ and $(f_h^m\circ f_v^n)(u_0)$ is measured in its ambient space and taken to be the magnitude of $2m+i2n\in\CC$.

\begin{lem}
Let $[\alpha]$ be homotopy class not in the kernel of $\varphi_1$. It will lift to an unclosed path terminating on $\varphi_1([\alpha])\cdot u_0$.
\end{lem}

\subsection{Monodromy Group Representation}
A \emph{rotational monodromy map} is a morphism from the first fundamental group of a surface to the \emph{monodromy group} acting on the unit tangent bundle via parallel transport of a vector along its path. Existence of a parallel transport map is a consequence of the metric connection on a surface with complex structure. The monodromy map is well defined when the \emph{holonomy group} is \emph{discrete}. When the \emph{relative holonomy} of contractible loops is trivial, the holonomy group is the set of attainable rotations of a unit vector upon returning to the base point after transporting it along a non-trivial loop. A \emph{flat surface} like $\bGs$ has trivial relative holonomy as its neighborhoods are isometric to the plane. Denote the \emph{unit tangent bundles} of $\bUs$ and $\bGs$ by $T^1\bUs$ and $T^1\bGs$. If $\alpha$ is a loop on $\bGs$, then $[\alpha]$ acts on a fixed vector $v_0\in\RR^2$ by rotation. In the case where $\alpha$ is \emph{not} a geodesic segment on the surface of length $L$, $v_0=\alpha'(0)\in\RR^2$ may differ from $\alpha'(L)$ by some rotation.

\begin{Def}
Define $\varphi_2:\pi_1(\bGs)\rightarrow SO(2,\ZZ)$ to be the group homomorphism where $a,b \mapsto I_2$, $c  \mapsto R$, and $d  \mapsto R^3$, where $I_2$ is the identity matrix and $R$ is a positive $90^\circ$ rotation matrix.
\end{Def}

@@

\begin{lem}
$\varphi_2:\pi_1(\bGs)\rightarrow \mathrm{SO}(2,\ZZ)$ is the rotational monodromy representation of $\bGs$.
\begin{proof}
From Figure 4 it's clear that $c,d$ will rotate a unit vector by $\pm 90^\circ$ by parallel transport along their paths. These values are determined by the cone angles of the singularities they loop around. The parallel transport of a vector along $a$ or $b$ would be trivial since the outer edges they pass through are glued by translation.
\end{proof}
\end{lem}

\begin{lem}
Any element in $\text{Deck}(\varphi_1^*)$ acting on $T^1\bUs$ has a trivial effect on holonomy.
\begin{proof}
Since $\bGs=\bUs/\ZZ^2$, the deck group action lifts naturally to an action on their tangent vector spaces and unit tangent bundles via pushforward of deck transformations $f_h,f_v$ on the bundle maps. i.e. $T^1\bGs=T^1\bUs/\ZZ^2$. This effect must be trivial since $\varphi_1(a)\sim f_h$ and $\varphi_1(b)\sim f_v$, and both homotopy classes $a,b$ have trivial holonomy.  \\
How do $f_h,f_v$ parallel transport a vector?\\
Well if $\varphi_1([a])\cdot u_0=f_h(u_0)$, then for 
\end{proof}
\end{lem}

\begin{thm}
Let $m=\varphi_2\vert_{\pi_1(\bUs)}$. Then $m$ is the rotational monodromy map of $\bUs$.
\begin{proof}

\begin{cor}
Let $\alpha:[0,1]\rightarrow\bGs$ be a loop on $\bGs$ with trivial holonomy and $\tilde{\alpha}$ its lift to $\bUs$. Then $\tilde{\alpha}$ has trivial holonomy as well.
\begin{proof}

\end{proof}
\end{cor}

\begin{Def}
Let $\bMs$ be the regular cover of $\bGs$ with fundamental group $\pi_1(\bMs)=\ker\varphi_2$.
\end{Def}
Showing that $\bMs$ is a degree four cover of $\bGs$ with trivial holonomy and deck group $\SO(2,\ZZ)$ follows immediately from its construction.
\begin{figure}[H]
\centering
\includegraphics[width=3in]{monogroup.png}
\label{fig:arbitrarylift}
\caption{Lift of paths $c,d$ to $\bMs$ (\compav{RELABEL EDGES})}
\end{figure}

@@

\subsection{Geodesics on $\bUs$}
The key connection made between surfaces $\bUs$ and $\btUs$ is that geodesics on $\bUs$ and $\btUs$ behave identically. Unless stated otherwise, a \emph{geodesic} is a \emph{unit-speed} locally distance minimizing curve on a smooth topological surface as a function parameterized by an element in $\RR$. A \emph{geodesic segment} is a restriction of the geodesic to some closed interval in $\RR$. A geodesic has an initial trajectory defined at time 0, represented by a unit vector $v_0\in\RR^2$. If the vector can be scaled by $k\in\RR$ such that $kv_0\in\cS$, the geodesic has a \emph{rational} initial trajectory. For example, the following is a periodic geodesic on the surface with initial direction $kv_0\in\cO$:

\begin{figure}[H]
\centering
\includegraphics[width=4in.]{closed2.png}
\caption{Periodic geodesic path on $\bUs$ modeled with sage-flatsurf.}
\label{fig:complicated}
\end{figure}

@@

\subsection{Translation Surface}
Consider the projection $\btU\rightarrow\mathbf{M}$ onto a quotient of the surface under its translational symmetries of each plane (see Fig 6). What we obtain is a genus 5 finite translation surface $\bM-$ also a branched cover of $\bG$. $\bMs$ was obtained as a cover of $\bGs$ from the rotational monodromy map on $\pi_1(\bGs)$, so $\bM$ is just its metric completion by reintroducing these vertices. There are four cone singularities on this surface, as opposed to the three on $\bG$, each of cone angle $6\pi$. Keeping track of $\btU$'s compass pointing \emph{true} north, the individual sections of $\bM$, labeled $I,II,III,$ and $IV$, are rotated in such a way to \emph{trivialize} \emph{holonomy} on arbitrary paths (Fig 8). Observe that $\bM$ is really a square-tiled/origami surface constructed out of unit squares (Fig 9).

@@

\noindent We recall some basic properties of $\ZZ^2$ covers of translation surfaces as they apply to $\bM$.

The set of all orientation preserving affine diffeomorphisms of $\bM$ form the group $\text{Aff}^+(\bM)$. The \emph{Veech group} of $\bM$ is the image of $\text{Aff}^+(\bM)$ in $SL(2,\RR)$ under the derivative map, $D:\text{Aff}^+(\bM)\rightarrow SL(2,\RR)$. We denote this group $V(\bM)$. It is a well-known fact that square-tiled surfaces like $\bM$ are \emph{Veech} and have Veech groups of finite index in $\SL(2,\ZZ)$ [Gutkin-Judge]. We use $V(M)$ and the $\text{Aff}^+(\bM)$-induced automorphism group, $\text{Aut}(\pi_1(\bM))$, as they act on geodesics and their homotopy classes. When we reconstructed $\bM$ as a staircase, we kept track of the outer edges of the surface that were identified with one another under the projection map defined on $\btU$. These now form cylinder core curves that are divided into homology classes $u,v,z$ (Fig 10). We denote the \emph{set of all core curve homology classes} by $\Gamma=\{[\gamma_i]~:~0\leq i \leq 11\}$.

\begin{figure}[H]
\includegraphics[width=3in]{homologyclass.png}
\centering
\caption{$\bM$'s cylinder core curves with u,v, and z homology class labels.}
\label{fig:homology}
\end{figure}

We assign $u,v$ classes integer weights according to what translational effect \emph{crossing} the curve would have on a loop being lifted to the cover $\btU$. For example, a path crossing $[\gamma_0]$ and making a positive angle relative to the direction $\gamma_0$ is moving in (right) corresponds to moving ``up" once on $\btU$ from whatever section in the fiber of $\bM$ under the projection map on $\btU$ the lifted path might have belonged to prior to intersecting with $\gamma_0$. We would then assign to $[\gamma_0]$ the weight $(0,+1)\in\ZZ^2$. Crossing a curve in the $z$ class really does not take a path to another element in the fiber.

@@

\subsection{Induced Automorphisms of $H_1(\bM,\QQ)$}
We look at some important automorphisms induced by Affine maps on $\bM$. Observe that $\bM$ has a uniform cylinder decomposition in both horizontal and vertical directions (Fig 15). We define the \emph{modulus}, $\mu$, of a cylinder to be the ratio of the cylinder's width to its circumference, $\frac{w}{c}$. The $\emph{Dehn-twist}$ of a cylinder is an affine diffeomorphism that skews the cylinder and preserves edges and vertices.

>>>>

\setlength{\topmargin}{1in} %top/bot margins
\setlength{\oddsidemargin}{\topmargin} %sidemargins

\setlength{\textheight}{11in} \setlength{\textwidth}{8.5in}
\setlength{\hoffset}{-1in} \setlength{\voffset}{-1in} \setlength{\evensidemargin}{\oddsidemargin} \addtolength{\textheight}{-2 \topmargin}\addtolength{\textwidth}{-2\oddsidemargin}
\setlength{\headheight}{0pt} \setlength{\headsep}{20pt} \setlength{\footskip}{20pt}
\addtolength{\textheight}{-\footskip} \addtolength{\textheight}{-\headheight} \addtolength{\textheight}{-\headsep}

@@

\subsection{Discussion of Results}
Our initial experiments supported the theory that there would be a strong correlation between a choice of initial trajectory angle and the dynamical properties of a geodesic traveling on the surface. [throw in some trajectories in the appendix] A surface composed of infinitely many cubes is endowed with a flat metric, i.e. every neighborhood is locally isometric to $\RR^2$. Via the parallel transport of a unit vector tangent to the surface in the direction of the geodesic, we view its image as a sequence of straight lines on the (smooth) faces of the surface that tightly wrap around its sharp edges (Fig 1).
\\
\begin{wrapfigure}{l}{2in.}
\includesvg{vectorplane}
\end{wrapfigure}
\noindent Let $u_0$ be a point on the surface that is contained in the interior of a face and consider a tangent unit vector $v\in\RR^3$ based at point $u_0$. Each face is parallel to a $2$-dimensional subspace of $\RR^3$ spanned by exactly two basis vectors, so one component of $v$ is always 0. Thus $v$ is identified with $v_0\in\RR^2$.
\\\\\\
Denote the Necker cube surface $\bUs$. The angle $v_0$ makes relative to a choice of basis is the \emph{initial trajectory angle} of a \emph{unit-speed geodesic} based at $u_0$ on the surface, $\Phi_t:\RR\rightarrow\bUs$. The initial trajectory angle is said to be $\emph{rational}$ if there is some $k\in\RR$ such that $kv_0$'s nonzero components are relatively prime integers. A rational initial trajectory angle in $\RR^2$ falls into one of two categories:
\begin{Def}
Let $v_0$ be a rational unit vector of the form $\frac{1}{k}(x,y)\in\RR^2$ with $x,y\in\mathbb{Z}$ and $k=\sqrt{x^2+y^2}\in\mathbb{R}$. We say $kv_0$ is an \textbf{odd-odd} vector if $x,y$ are relatively prime and both odd. We denote the \textbf{set of all odd-odd directions} by $\mathcal{O}$. We say that $kv_0$ is an \textbf{even-odd} vector if $x,y$ are relatively prime and of opposite parity. We denote the \textbf{set of all even-odd directions} by $\mathcal{E}$.
\end{Def}
\noindent Let $\cS=\cO\sqcup\cE$ be the disjoint union of the two. This paper aims to depict the relationship between initial trajectory angles and dynamical properties of a geodesic on the surface. Recall that a geodesic on the surface is \emph{periodic} if there exists some $T>0$ such that $\Phi(t+T)=\Phi(t)$ for all $t\in\RR$ and a geodesic is \emph{drift-periodic} if there exists some $T'>0$ and non-trivial translation, $p$, such that $\Phi(t+T')=p(\Phi(t))$ for all $t\in\RR$. In either case we call this value the \emph{period} of $\Phi$ if it is the \emph{smallest} possible value that satisfies this definition. We additionally require that at no point does $\Phi$ encounter a cone singularity of the surface. Otherwise the geodesic behavior is undefined. We call $\Phi$ \emph{non-singular} if this holds.

@@

\begin{thm*}
(Dynamical Classification of Geodesics on the Necker Cube Surface) \\Let $\Phi_t:\RR\rightarrow\bUs$ be a non-singular unit-speed geodesic on the Necker cube surface such that $\Phi(0)=u_0$ and $\Phi'(0)=v\in\RR^3$. Let $v_0\RR^2$ be a distance-preserving projection of $v$ onto the plane tangent to the the surface at $u_0$. Let $\overline{u}_t\in\RR^3$ be $\Phi_t$'s position at time $t$. Let $k\in\RR$ such that $kv_0\in\cS$. Then the following is true:
\begin{enumerate}[label=(\roman*)]
\item $\Phi$ is periodic if $kv_0\in\cO$.
\item Suppose $\Phi$ is periodic. If $M (kv_0)=(1,1)$ then $\Phi$ has period $T=6\sqrt{2}||kv_0||$.
\item $\Phi$ is drift-periodic if $v_0\in\cE$. 
\item Suppose $\Phi$ is drift-periodic. If $M (kv_0)=(1,0)$ or $M (kv_0)=(0,1)$ then $\Phi$ has period $T=2||kv_0||$, and $d_E(\overline{u}_t,\overline{u}_{t+T})=||kv_0||\sqrt{2}$.
\end{enumerate}
\end{thm*}
\compav{Remember to prove that such an M exists.}

@@

We will call this ``half-cube" $\bG$. It is a piecewise smooth surface homeomorphic to the two-torus, $\TT^2$. Every vertex has a cone angle of either $3\pi$ or $\frac{3\pi}{2}$. Let $\Sigma_{\bG}\subset\bG$ be the set of these three vertices. These are conical singularities of the surface. We denote the surface punctured at these points by $\bG\backslash\Sigma_{\bG} = \bGs$. The paths labeled $a,b,c,d$ are independent of one another and span $\bGs$'s fundamental group, $\pi_1(\bGs)\cong \<a,b,c,d \>$, the free group of four generators. The unpunctured Necker cube surface $\bU$ is embedded in $\RR^3$ by tiling along $\bG$'s outer edges:

\begin{figure}[H]
\centering
\includesvg[width=4in]{neckercubepathsurface}
\end{figure}
From this diagram we see that $\bU$ is a universal cover of $\bG$. In this paper we are studying geodesic behavior on $\bU$ and in order to do so, it is necessary to determine how the parallel transport of a unit vector around a cone singularity acts on the unit tangent bundle of the surface. But to study a given path's rotational holonomy, we require that these loops (see $c,d$ in Fig 3) aren't contractible through the cube's sharp corners. To that effect we puncture the surface at these points.  Denote the countably infinite set of singularities of $\bU$ by $\Sigma_{\bU}$. The \emph{Necker cube surface} is the punctured surface $\bUs=\bU\backslash\Sigma_{\bU}$. Observe from the previous diagram that $\pi_1(\bUs)$ is the kernel of the following group homomorphism:

\begin{Def}
Let $\varphi_1:\pi_1(\bGs)\rightarrow \ZZ^2$ be the group homomorphism given as $c,d\mapsto(0,0)$, $a\mapsto(1,0)$, and $b\mapsto(0,1)$.
\end{Def}

The cover of $\bGs$ with fundamental group $\ker\varphi_1$ is the space $\bUs=\tilde{\bGs}/\ker\varphi_1$ where $\ker\varphi_1$ acts on $\bGs$'s universal cover, $\tilde{\bGs}$, by deck transformations. We say $\Delta_{\varphi_1}=\pi_1(\bGs)/\pi_1(\bUs)\cong\ZZ^2$ is the \emph{deck group} of the covering map $\varphi_1^*:\bUs\rightarrow\bGs$. Now $\bGs$ is unfolded and flattened to better facilitate the study of its rotational holonomy. Apply the following cutting and gluing operations to the surface: \\
\begin{figure}[H]
\centering
\includesvg[width=3in]{LTorus}
\end{figure}
\noindent This L-shaped torus is then taken apart and recovered as $2\times 2$ torus missing a unit square with the following identifications:\\
\begin{figure}[H]
\centering
\includesvg[width=2in]{BranchTorus}
\caption{$\bGs$ and images of paths after reconstruction. The outermost vertex is \emph{not} a puncture.}
\end{figure}

@@

\compav{Include diagram of P}

\noindent Let $2m+i2n\in\CC$ be the center of a removed square with $m,n\in\ZZ$. Now let $\zeta_{m,n}(t)=(2m+t)+i(2n-\frac{1}{2})$ and 
$\xi_{m,n}(t)=(2m+t)+i(2n+\frac{1}{2})$ be paths on $\mathbf{P}$ that run along the top and bottom edges of the squares, parameterized by $t\in (-\frac{1}{2},\frac{1}{2})$ and $(m,n)\in\ZZ^2$. $\bUs$ is recovered as a topological quotient $\mathbf{P}/\sim$, where $\sim$ is the minimal relation on $\mathbf{P}$:

\begin{equation}
\begin{split}
\zeta(t)\sim e^{i\frac{\pi}{2}}~\overline{\zeta(t)-(2m+i2n)}\\
\eta(t)\sim e^{i\frac{\pi}{2}}~\overline{\eta(t)-(2m+i2n)}
\label{eq:rel2}
\end{split}
\end{equation}
The translations, $\pm 2,\pm 2i$, on $\mathbf{P}$ are isometries that preserve these identifications and induce isometries of $\bUs$. We denote the isometry induced by $+2$ as $f_h$, for horizontal, and the isometry induced by $+2i$ as $f_v$, for vertical. The group generated by these horizontal and vertical translations form the \emph{group of Deck transformations}, $\text{Deck}(\varphi_1^*)$, isomorphic to $\ZZ^2$:

@@

One can see that the Let $x_0\in\bGs$ and fix a point in the fiber,  $u_0\in\varphi_1^*{-1}(x_0)\subset\bUs$. The action of $\ZZ^2$ on the fiber over $x_0$ is given by $(m,n)\cdot u_0=(f_h^m\circ f_v^n)(u_0)$.

\compav{Discuss:}
The distance between $u_0$ and $(f_h^m\circ f_v^n)(u_0)$ is measured in its ambient space and taken to be the magnitude of $2m+i2n\in\CC$.

\begin{lem}
Let $\alpha:[0,1]\rightarrow \bGs$ be a loop on the surface. Let $[\alpha]\in\pi_1(\bGs)$ be its homotopy class representative. Denote its lift to $\bUs$ based at $u_0$ by $\tilde{\alpha}$. Then $\tilde{\alpha}(1)=\varphi([\alpha])\cdot\tilde{\alpha(0)}$.
\end{lem}

\subsection{Monodromy Group Representation}
A \emph{rotational monodromy map} is a morphism from the first fundamental group of a surface to the \emph{monodromy group} acting on the unit tangent space via parallel transport of a vector along its path. Existence of a parallel transport map is a consequence of the metric connection on a surface with complex structure.
\compav{Discuss:}\\
The monodromy map is well defined when the \emph{holonomy group} is \emph{discrete}. When the \emph{relative holonomy} of contractible loops is trivial, the holonomy group is the set of attainable rotations of a unit vector upon returning to the base point after transporting it along a non-trivial loop. A \emph{flat surface} like $\bGs$ has trivial relative holonomy as its neighborhoods are isometric to the plane. Denote the \emph{unit tangent bundles} of $\bUs$ and $\bGs$ by $T^1\bUs$ and $T^1\bGs$. If $\alpha$ is a loop on $\bGs$, then $[\alpha]$ acts on a fixed vector $v_0\in\RR^2$ by rotation. In the case where $\alpha$ is \emph{not} a geodesic segment on the surface of length $L$, $v_0=\alpha'(0)\in\RR^2$ may differ from $\alpha'(L)$ by some rotation.

\begin{Def}
Define $\varphi_2:\pi_1(\bGs)\rightarrow SO(2,\ZZ)$ to be the group homomorphism where $a,b \mapsto I_2$, $c  \mapsto R$, and $d  \mapsto R^3$, where $I_2=\begin{bmatrix}1 & 0\\0 & 1\end{bmatrix}$ and $R=\begin{bmatrix}0 & -1 \\1 & 0\end{bmatrix}$.
\end{Def}

@@

\begin{thm}
$\varphi_2:\pi_1(\bGs)\rightarrow \mathrm{SO}(2,\ZZ)$ is the rotational monodromy representation of $\bGs$.
\begin{proof}
From Figure 4 it's clear that $c,d$ will rotate a unit vector by $\pm 90^\circ$ by parallel transport along their paths. These values are determined by the cone angles of the singularities they loop around. The parallel transport of a vector along $a$ or $b$ would be trivial since the outer edges they pass through are glued by translation.
\end{proof}
\end{thm}

\compav{Discuss:}
\begin{lem}
Any element in $\text{Deck}(\varphi_1^*)$ acting on $T^1\bUs$ has a trivial effect on holonomy.
\begin{proof}
Since $\bGs=\bUs/\ZZ^2$, the deck group action lifts naturally to an action on their tangent vector spaces and unit tangent bundles via pushforward of deck transformations $f_h,f_v$ on the bundle maps. i.e. $T^1\bGs=T^1\bUs/\ZZ^2$. This effect must be trivial since $\varphi_1(a)\sim f_h$ and $\varphi_1(b)\sim f_v$, and both homotopy classes $a,b$ have trivial holonomy.  \\
How do $f_h,f_v$ parallel transport a vector?\\
Well if $\varphi_1([a])\cdot u_0=f_h(u_0)$, then for 
\end{proof}
\end{lem}

A ramified cover of $\bGs$ is obtained by cyclically gluing 4 copies of neighborhoods surrounding its cone singularities. This cover is ramified over three singularities of cone angle $\frac{3\pi}{2}$ and $3\pi$, so 4 is the minimal number of copies needed to give every singularity in the cover a cone angle that is a multiple of $2\pi$. We call this cover $\bMs$.

\begin{figure}[H]
\centering
\includegraphics[width=3in]{monogroup.png}
\label{fig:arbitrarylift}
\caption{Lift of paths $c,d$ to $\bMs$ (\compav{RELABEL EDGES})}
\end{figure}

\begin{Def}
Let $\bMs$ be the regular cover of $\bGs$ with fundamental group $\pi_1(\bMs)=\ker\varphi_2$.
\end{Def}
\compav{Mark the puncutres}\\
That $\bMs$ is a degree four cover of $\bGs$ with trivial rotational holonomy and deck group $\SO(2,\ZZ)$ follows immediately from its construction. The metric completion of $\bMs$ is the compact translation surface $\bM$ obtained by reinserting its cone singularities. A \emph{translation surface} is a topological surface with a discrete set of cone singularities and a maximal translation atlas on the surface with these singularities removed. Neighborhoods of the surface are naturally isometric to $\CC$, and a metric on the translation surface is obtained by pulling back the Euclidean metric in $\CC$.

@@

\subsection{Geodesics on $\bUs$}
The key connection made between surfaces $\bUs$ and $\btUs$ is that geodesics on $\bUs$ and $\btUs$ behave identically. A \emph{geodesic} is a \emph{unit-speed} locally distance minimizing curve on a smooth surface as a continuous function parameterized by an element in $\RR$. A \emph{geodesic segment} is a restriction of the geodesic to some closed interval in $\RR$. A geodesic has derivative defined at time 0, represented by a unit vector $v_0\in\RR^2$. If the vector can be scaled by $k\in\RR$ such that $kv_0\in\cS$ (defined in the introduction), the geodesic has a \emph{rational} initial trajectory angle. For example, the following is a periodic geodesic on the surface with initial direction $kv_0\in\cO$:
\\\compav{find this direction}
\begin{figure}[H]
\centering
\includegraphics[width=4in.]{closed2.png}
\caption{Periodic geodesic path on $\bUs$ modeled with sage-flatsurf.}
\label{fig:complicated}
\end{figure}

@@

\subsection{Translation Surface}
The surface $\bMs$ was obtained as a cover of $\bGs$ from the rotational monodromy map on $\pi_1(\bGs)$, and $\bM$ is its metric completion as well as a branch cover of $\bGs$. There are four cone singularities on this surface, as opposed to the three on $\bG$, each of cone angle $6\pi$. The individual sections of $\bM$, labeled $I,II,III,$ and $IV$, are rotated in such a way to trivialize holonomy on arbitrary paths (Fig 8). Though not immediately apparent, this surface can be realized as a finite ``staircase" surface. (Fig 9).

@@

\noindent The set of all orientation preserving affine diffeomorphisms of $\bM$ form the group $\text{Aff}^+(\bM)$. The \emph{Veech group} of $\bM$ is the image of $\text{Aff}^+(\bM)$ in $SL(2,\RR)$ under the derivative map, $D:\text{Aff}^+(\bM)\rightarrow SL(2,\RR)$. We denote this group $V(\bM)$. We have two transverse cylinder decompositions of this surface in vertical and horizontal directions. Let $\mathfrak{H}=\{\delta_A,\delta_C,~\cdots,~\delta_K \}$ and $\mathfrak{V}=\{\epsilon_B,\epsilon_D,~\cdots,~\epsilon_{L/M} \}$ be the set of maximal cylinder decompositions of $\bM$ in the horizontal and vertical directions respectively. Their subscripts denote which edges the cylinder waist curves pass through. These cylinders are uniform in size, making $\bM$ a particularly symmetric surface. A convenient way to index the set of waist curves is in the following manner:

\begin{figure}[H]
\includegraphics[width=3in]{homologyclass.png}
\centering
\caption{$\bM$'s core curves with u,v, and z homology class labels.}
\label{fig:homology}
\end{figure}

These curves associate an even indexed $\gamma$ with a horizontal cylinder, and an odd indexed $\gamma$ with a vertical cylinder. Let $\Gamma=\{[\gamma_i]~:~0\leq i \leq 11\}$ be the set of homology classes of these curves. We will want to work with $\bM's$ homology group, $H_1(\bM,\QQ)$.

\begin{lem}
$\Gamma$ generates $H_1(\bM,\QQ)$.
\begin{proof}
One can show this is true by constructing a $12\times 12$ intersection matrix and calculating its rank. The genus of $\bM$ is 5, so $H_1(\bM,\QQ)\cong\QQ^{10}$. The rank of such a matrix is 10, which agrees with the dimension of its homology.
\end{proof}
\end{lem}

\noindent $\Gamma$ is further divided into three classes:
\begin{Def}
The homology classes $u,v,z$ are given as the following weighted sums of elements of $\Gamma$:
\begin{align*}
u &= -[\gamma_2] +[\gamma_5] + [\gamma_8] - [\gamma_{11}],\\
v &= +[\gamma_0] -[\gamma_3] -[\gamma_6] +[\gamma_9],\\
z &= +[\gamma_1] +[\gamma_4]-[\gamma_7]-[\gamma_{10}].
\end{align*}
\end{Def}

The curves composing $u$ and $v$ are linearly independent. By observation, these curves are weighted according to how a loop $\alpha:[0,1]\rightarrow\bM$ would lift to a path on $\btU$ by its intersections with these waist curves. For example, a path crossing $[\gamma_0]$ and making a positive angle relative to $\gamma_0$'s direction corresponds to moving ``up" a section on $\btU$. Because we consider up to be positive, $[\gamma_0]$ is signed positively, and we would give it the weight $(0,+1)\in\ZZ^2$.

@@

\subsection{Induced Automorphisms of $H_1(\bM,\QQ)$}
We have previously remarked on the uniformity of $\bM$'s cylinder decompositions, $\mathfrak{H},~\mathfrak{V}$, but have yet to show the correlation between these cylinders and $\text{Aff}^+(\bM)$. We say the \emph{modulus}, $\mu_i$, of the cylinder containing the $i^{th}$ waist curve to be the ratio of its width to its circumference, $\frac{w_i}{c_i}$. The $\emph{Dehn-twist}$ of a cylinder in either set is an affine diffeomorphism that skews the cylinder and preserves edges and vertices.
